---
title: "Hands-on Exercise 12"

execute:
  warning: false
  echo: true
  eval: true
  
date: "`r Sys.Date()`"
---

# R packages

```{r}
pacman::p_load(tmap, SpatialAcc, sf, 
               ggstatsplot, reshape2,
               tidyverse)
```

# Geospatial data wrangling

## Import data

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL")
```

```{r}
hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons") 
```

```{r}
eldercare <- st_read(dsn = "data/geospatial", layer = "ELDERCARE") 
```

## Update CRS info

```{r}
mpsz <- st_transform(mpsz, 3414)
eldercare <- st_transform(eldercare, 3414)
hexagons <- st_transform(hexagons, 3414)
```

## Cleaning and updating attribute fields of the geospatial data

There are many redundant fields in the data tables of both `eldercare` and `hexagons`. The code chunks below will be used to exclude those redundant fields. At the same time, a new field called `demand` and a new field called `capacity` will be added into the data table of `hexagons` and `eldercare` sf data frame respectively. Both fields are derive using *mutate()* of **dplyr** package.

```{r}
eldercare <- eldercare %>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)
```

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

# Aspatial data wrangling

## Import data

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```

## Tidying it up

::: callout-note
The rows and columns align with the exit/entry/total costs in the OD matrix
:::

The code chunk below uses *spread()* of **tidyr** package is used to transform the O-D matrix from a thin format into a fat format.

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  spread(destination_id, total_cost)%>%
  select(c(-c('origin_id')))
```

Currently, the distance is measured in metre because SVY21 projected coordinate system is used. The code chunk below will be used to convert the unit f measurement from metre to kilometre.

```{r}
distmat_km <- as.matrix(distmat/1000)
```

# Modelling and Visualising Accessibility using Hansen Method

## Computing Hansen’s accessibility

Now, we ready to compute Hansen’s accessibility by using *ac()* of [**SpatialAcc**](https://cran.r-project.org/web/packages/SpatialAcc/) package. Before getting started, you are encourage to read the [arguments](https://cran.r-project.org/web/packages/SpatialAcc/SpatialAcc.pdf) of the function at least once in order to ensure that the required inputs are available.

The code chunk below calculates Hansen’s accessibility using *ac()* of **SpatialAcc** and *data.frame()* is used to save the output in a data frame called `acc_Handsen`.

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 2, 
                            family = "Hansen"))
```

The default field name is very messy, we will rename it to `accHansen` by using the code chunk below.

```{r}
colnames(acc_Hansen) <- "accHansen"
```

Next, we will convert the data table into tibble format by using the code chunk below.

```{r}
acc_Hansen <- tbl_df(acc_Hansen)
```

Lastly, *bind_cols()* of dplyr will be used to join the acc_Hansen tibble data frame with the hexagons simple feature data frame. The output is called `hexagon_Hansen`.

```{r}
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

## Visualising Hansen’s accessibility

### Extracting map extend

```{r}
mapex <- st_bbox(hexagons)
```

```{r}
tmap_mode("plot")
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_fill(col = "accHansen",
          n = 10,
          style = "quantile",
          border.col = "black",
          border.lwd = 1) +
tm_shape(eldercare) +
  tm_symbols(size = 0.1) +
  tm_layout(main.title = "Accessibility to eldercare: Hansen method",
            main.title.position = "center",
            main.title.size = 2,
            legend.outside = FALSE,
            legend.height = 0.45, 
            legend.width = 3.0,
            legend.format = list(digits = 6),
            legend.position = c("right", "top"),
            frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.5)
```

### Statistical graphic visualisation

In this section, we are going to compare the distribution of Hansen’s accessibility values by URA Planning Region.

Firstly, we need to add the planning region field into *haxegon_Hansen* simple feature data frame by using the code chunk below.

```{r}
hexagon_Hansen <- st_join(hexagon_Hansen, mpsz, 
                          join = st_intersects)
```

```{r}
ggplot(data=hexagon_Hansen, 
       aes(y = log(accHansen), 
           x= REGION_N)) +
  geom_boxplot() +
  geom_point(stat="summary", 
             fun.y="mean", 
             colour ="red", 
             size=2)
```

::: callout-note
Use a HISTOGRAM to show skewness in distribution instead of a boxplot
:::
