---
title: "Hands-on Exercise 2: Thematic Mapping and GeoVisualisation with R"

execute:
  warning: false
date: "`r Sys.Date()`"
---

In this exercise, we learn how to plot and visualize choropleth maps using an R package called \*\*tmap\*\* package.

## Installing and loading R packages

The packages that are used in this hands on exercise are:

-   [**tmap**](https://cran.r-project.org/web/packages/tmap/)

-   [**readr**](https://readr.tidyverse.org/) for importing delimited text file

-   [**tidyr**](https://tidyr.tidyverse.org/) for tidying data

-   [**dplyr**](https://dplyr.tidyverse.org/) for wrangling data

-   [**sf**](https://cran.r-project.org/web/packages/sf/) for handling geospatial data

We use the following to install and load the packages in Rstudio:

```{r}
#| eval: false
pacman::p_load(sf, tmap, tidyverse)
```

(Note that readr, tidyr and dplyr are part of tidyverse package)

## Importing the data

Two data sets are needed to create the choropleth map. They are:

-   Master Plan 2014 Subzone Boundary (Web) in ESRI format from [data.gov.sg](https://beta.data.gov.sg/)

-   Singapore Residents by Planning Area/Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 in csv format from the [Department of Statistics, Singapore](https://www.singstat.gov.sg/)

## Importing Data into R
### Geospatial data
We will use *st_read* to import the geospatial data into R.
```{r}
mpsz <- st_read(dsn = "chap02/data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```


### Aspatial data